---
title: "Reanalysis of Nitrogen Depletion RNA-seq in *Chlamydomonas reinhardtii*"
subtitle: "Reanalysis of Miller et al. 2010 using DESeq2"
author: "Matthew Francoeur"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# Script: deseq2_analysis.Rmd
# Description: Differential expression analysis of *C. reinhardtii* nitrogen
#              deprivation RNA-seq data using DESeq2
# Author: Matthew Francoeur
# Date: 2026

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)

library(DESeq2)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
``` 
## Introduction
This document outlines the differential expression analysis portion of the *C. reinhardtii* nitrogen deprivation reanalysis study. Here the count matrices generated by StringTie will be assessed sample quality and consistency before being passed to DESeq2 for differential expression analysis. Results will be summarized and exported for downstream analysis. For full context on this reanalysis study, see the README.  

## Data Import and Inspection
In this section, the gene count matrix is imported and its basic attributes inspected

```{r load-data}
# Load count matrix
count_matrix <- read.csv("~/Desktop/Projects/Chlamydomonas_rnaseq/results/counts/gene_count_matrix.csv", 
                   row.names = 1)

# Inspect dimensions
dim(count_matrix)

# Preview first few rows
head(count_matrix)
```
The count matrix contains 6 samples with 16,661 genes. The first 6 rows of this matrix are shown. Some genes do not have a known gene name in the annotation, which are identified by their MSTRG ID only.  

## Exploratory Analysis
### Coldata Construction
```{r coldata construction}
# Make data frame
coldata_df <- data.frame(
  sample = c("SRR066643", "SRR066644", "SRR066645", "SRR066646", "SRR066647", "SRR066648"), 
  condition = c("Replete", "Replete", "Replete", "Deprived", "Deprived", "Deprived")
)
row.names(coldata_df) <- coldata_df$sample
coldata_df
```
A coldata data frame is defined, mapping each sample to its experimental condition for use by DESeq2

### Build DeSeq2 Object
```{r Build DeSeq2 object}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata_df,
  design = ~condition
)
```
### PCA Plot
```{r PCA Plot}
# Apply variance stabilizing transformation
vsd <- vst(dds, blind = TRUE) 

#Create PCA plot
plotPCA(vsd, intgroup = "condition") + theme_minimal()
```
A variance stabilizing transformation is applied to the DESeq data set to stabilize gene expression level variation within each sample. A PCA plot is then created from the transformed data and colored by treatment condition. From the resulting plot, it is clear that the samples separate cleanly by treatment group along PC1, which accounts for 91% of the variation in the data set. This indicates that the nitrogen deprivation condition created a robust transcriptional response that will be picked up by DESeq2. PC2 accounts for 4% of the variation in the data set, with the replete condition showing notable separation along this axis. This, as well as greater variation along PC1, indicates potential technical variability among the replete condition, though this does not undermine the analysis given the strong distinction between sample conditions.  

### Sample Distance Heatmap
